---
title: "nlmixr2extra events"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{nlmixr2extra events}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
## The examples need an rxode2 with the event bus; without it every chunk is
## shown but not run.
has_bus <- exists("rxEventListen", envir = asNamespace("rxode2"), inherits = FALSE)
knitr::opts_chunk$set(eval = has_bus)
```

```{r no-bus, eval = !has_bus, echo = FALSE, results = "asis"}
cat("> **Note:** the installed `rxode2` has no event bus",
    "(`rxode2::rxEventListen()`), so the code in this article is shown but",
    "not run.\n")
```

# Introduction

`rxode2` provides a small event bus: packages *emit* an event when a
user-facing operation finishes, and any package can *listen* for it with
`rxode2::rxEventListen()`.  The main consumer is
[nlmixr2log](https://github.com/nlmixr2/nlmixr2log), which records every fit,
simulation and result as a run, but a listener can be any function.  See
`?rxode2::rxEventListen` for the bus itself.

The key rule of the bus is that events are only delivered from the
**outermost** operation.  Every emitting function enters a shared operation
scope, and an event is dropped when it is emitted inside one.  That matters a
great deal for `nlmixr2extra`: its drivers (`bootstrapFit()`, `multistart()`,
`covarSearchAuto()`, ...) run tens or hundreds of `nlmixr2()` fits and
`rxSolve()` calls internally.  Each driver enters the scope, so its internal
fits and solves are silent, and on the way out it emits **one** event that
describes the whole call.  A logger then stores one entry per driver call
instead of one run per internal fit.

Everything here is a no-op when the installed `rxode2` has no event bus, and
nothing is emitted when no listener is registered.

# Which event a driver emits

On exit, a driver looks at what it returned and picks one event:

* the input fit itself, changed in place, gives **`fitUpdate`** -- a new
  version of that fit's run;
* a different fit gives **`fitComplete`** with `source` set to the driver's
  kind -- a new run linked to the input fit (`object`);
* anything else (a data frame, a list of results, a plot) gives one
  **`fitResult`** carrying a fit-free summary of the result.

When a result holds a best fit (`multistart()`'s `$best`), that fit is also
announced with `fitComplete`, linked to the input fit, so it can be kept as
its own run.

| Driver                          | Event(s)                                    | `what` / `source` / `kind` |
|:--------------------------------|:--------------------------------------------|:---------------------------|
| `bootstrapFit()` (default `returnType = "model"`) | `fitUpdate`                | `bootstrap`                |
| `bootstrapFit(returnType = "fitList"/"modelList")` | `fitResult`               | `bootstrap`                |
| `preconditionFit()`             | `fitUpdate` (always)                        | `covariance`               |
| `multistart(fit)`               | `fitResult`, then `fitComplete` for `$best` | `multistart`               |
| `multistart(model, data)`       | `fitComplete` for `$best`, then `fitResult` | `multistart`               |
| `covarSearchAuto()`             | `fitResult`                                 | `covarSearch`              |
| `resSearch()`                   | `fitResult`                                 | `resSearch`                |
| `iivSearch()`                   | `fitResult`                                 | `iivSearch`                |
| `rerunTopN()`                   | nothing (refits are silent)                 | --                         |
| `linearize()`                   | `fitComplete`                               | `linearize`                |
| `profile()` (fit method)        | `fitResult`                                 | `profile`                  |
| `profileFixed()`                | `fitResult`                                 | `profile`                  |
| `profileLlp()`                  | `fitResult`                                 | `profile`                  |
| `lassoCoefficients()`           | `fitResult`                                 | `lasso`                    |
| `adaptivelassoCoefficients()`   | `fitResult`                                 | `lasso`                    |
| `adjustedlassoCoefficients()`   | `fitResult`                                 | `lasso`                    |
| `regularmodel()`                | `fitComplete`                               | `lasso`                    |
| `bootplot()`                    | `fitResult` (the plot)                      | `bootplot`                 |

The kind is stored in `what` for `fitUpdate`, in `source` for `fitComplete`
and in `kind` for `fitResult`.

# Payloads

Listeners are called as `fun(event, ...)`: the event name, then the payload
fields as named arguments.

## `fitUpdate`

| Field      | Value                                                     |
|:-----------|:----------------------------------------------------------|
| `fit`      | the input fit, as changed by the driver                   |
| `original` | the same fit (it was changed in place)                    |
| `name`     | `NULL`                                                    |
| `what`     | the driver's kind, e.g. `"bootstrap"` or `"covariance"`   |
| `inPlace`  | `TRUE`                                                    |
| `call`     | the driver call                                           |

## `fitComplete`

| Field     | Value                                                                   |
|:----------|:------------------------------------------------------------------------|
| `fit`     | the new fit (`linearize()`, `regularmodel()`, or a result's best fit)   |
| `object`  | the input fit it came from; `NULL` when the driver started from a model |
| `call`    | the driver call                                                         |
| `objName` | `NULL`                                                                  |
| `source`  | the driver's kind, e.g. `"linearize"`                                   |

## `fitResult`

| Field    | Value                                                                  |
|:---------|:-----------------------------------------------------------------------|
| `fit`    | the input fit (or the best fit when the driver started from a model)   |
| `result` | a fit-free summary of what the driver returned (see below)             |
| `kind`   | the driver's kind, e.g. `"profile"`                                    |
| `call`   | the driver call                                                        |

The `result` is built so that a logger can store it cheaply:

* every fit inside the result is replaced by
  `list(OBJF = <objective function>, parFixedDf = <parameter table>)`, read
  from the fit without computing anything;
* data frames and `ggplot` objects are kept as they are;
* functions and environments are dropped;
* lists are summarized up to four levels deep, and the original class of a
  classed list is kept in the `"nlmixr2extraClass"` attribute.

The `call` of every event is normalized by `rxode2`: its head is the driver's
name, and large values inlined in it are replaced by markers, so a recorded
call never embeds a data set or a fit.

# What is not emitted

* **Internal work.** The fits, refits and solves a driver runs are inside its
  scope and emit nothing -- not the bootstrap fits, not the multistart
  candidates, not the profile steps, not the covariate search candidates.
  This includes `bootstrapFit()` when it is called by `bootplot()`.
* **Nested drivers.** A driver called from inside another driver (or from
  inside any other emitting operation, such as `nlmixr2()`) is itself
  silent; only the outermost call reports.
* **Errors.** When a driver fails, it leaves the scope and emits nothing.
* **Non-fit inputs.** A driver whose input is not a fit emits only when its
  result holds a best fit (as `multistart(model, data)` does).  So
  `rerunTopN()`, whose input is an `iivSearch()` result, emits nothing; its
  refits are simply kept silent.
* **A best fit that is the input.** When a result's best fit is the input
  fit itself, only the `fitResult` is emitted.
* **Unsummarizable results.** If summarizing the result fails, the driver
  leaves the scope without an event rather than breaking the call.

# Listening

A listener is a function `function(event, ...)`.  This one records the event
name and the driver's kind:

```{r listen, message = FALSE}
library(nlmixr2extra)

events <- list()
rxode2::rxEventListen("demo", function(event, ...) {
  p <- list(...)
  kind <- if (!is.null(p$what)) p$what else if (!is.null(p$source)) p$source else p$kind
  events[[length(events) + 1L]] <<- data.frame(event = event, kind = kind)
})
```

A fit to drive.  It is run inside `rxode2::rxEventScope()` here so that only
the drivers below are recorded (otherwise `nlmixr2()` would report this fit
with its own `fitComplete`, when `nlmixr2est` has the event hooks):

```{r fit, message = FALSE, warning = FALSE}
one.compartment <- function() {
  ini({
    tka <- log(1.57)
    tcl <- log(2.72)
    tv <- log(31.5)
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v ~ 0.1
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v <- exp(tv + eta.v)
    linCmt() ~ add(add.sd)
  })
}

fit <- rxode2::rxEventScope(
  nlmixr2(one.compartment, nlmixr2data::theo_sd, est = "focei",
          control = list(print = 0))
)
```

Now call a few drivers.  Each runs many fits internally, but reports once (or
twice, for `multistart()`'s best start):

```{r drivers, message = FALSE, warning = FALSE, results = "hide"}
preconditionFit(fit)
lin <- linearize(fit)
prof <- profile(fit, which = data.frame(tka = log(c(1.4, 1.6))), method = "fixed")
ms <- multistart(fit, control = list(n = 2, spread = 0.3, screen = "none",
                                     cacheDir = NA, refitBest = FALSE))
```

```{r show}
do.call(rbind, events)
```

The heavier drivers report the same way; for example, this records a single
`fitUpdate` (kind `bootstrap`) for the 100 bootstrap fits:

```{r bootstrap, eval = FALSE}
bootstrapFit(fit, nboot = 100)
```

When you are done, remove the listener:

```{r unlisten}
rxode2::rxEventUnlisten("demo")
```

# Full logging

For a complete log of fits, simulations and driver results -- with each
driver call stored as one entry linked to the fit it started from -- use
[nlmixr2log](https://github.com/nlmixr2/nlmixr2log), which listens to these
events for you.
