---
title: "Helpers for writing modeling reports"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Helpers for writing modeling reports}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The `nlmixr2` family of packages makes creating reports easier. When you fit
several candidate models, a few common tasks come up over and over again:

* deciding whether a fit can be trusted (did any parameter land on a boundary?),
* choosing the "best" model by an information criterion, and
* summarizing every model that was tested in a single table.

`nlmixr2extra` provides three helpers for these tasks:

* `isBoundaryFit()` reports whether a fit has a parameter at its boundary.
* `getMinAICFit()` returns the fit with the lowest AIC, optionally excluding
  boundary fits and silently ignoring fits that errored.
* `listModelsTested()` builds a report-ready table of every model tested with
  its AIC and change from the minimum AIC (dAIC).

The functions work for a single model or for many models at once.

## Setup

Start with your data and create the candidate models you want to compare. Here
the data follow a step change (a low value at `conc == 0` and a higher value for
any positive concentration), and we fit three competing structural models.

```{r setup}
library(nlmixr2est)
library(nlmixr2extra)

# Start with your data
d_noec50 <-
  data.frame(
    conc = c(rep(0, 10), rep(1:20, each = 10)),
    DV = c(rnorm(n = 10, mean = 1, sd = 1e-5), rnorm(n = 200, mean = 5, sd = 1e-5)),
    TIME = 0
  )

# An Emax model.  Because the data are a step change, ec50 is pushed to its
# lower boundary (0), which makes this fit unreliable.
modEmax <- function() {
  ini({
    e0 = 1
    emax = 5
    ec50 = c(0, 1.1)
    addSd = 0.5
  })
  model({
    effect <- e0 + emax*conc/(ec50 + conc)
    effect ~ add(addSd)
  })
}

# A step-change model
modStep <- function() {
  ini({
    e0 = 1
    emax = 5
    addSd = 1e-5
  })
  model({
    effect <- e0 + emax*(conc > 0)
    effect ~ add(addSd)
  })
}

# A linear model
modLinear <- function() {
  ini({
    e0 = 1
    slope = 5
    addSd = 1
  })
  model({
    effect <- e0 + slope*conc
    effect ~ add(addSd)
  })
}

# Fit the models
fitEmaxBoundaryIssue <- nlmixr2est::nlmixr2(modEmax, data = d_noec50, est = "focei", control = list(print = 0))
fitStep <- nlmixr2est::nlmixr2(modStep, data = d_noec50, est = "focei", control = list(print = 0))
fitLinear <- nlmixr2est::nlmixr2(modLinear, data = d_noec50, est = "focei", control = list(print = 0))
```

# Detecting boundary issues

A model whose estimate sits on a parameter boundary usually should not be
trusted or selected. `isBoundaryFit()` returns `TRUE` when a fit has a parameter
at its boundary and `FALSE` otherwise (including for objects that are not
`nlmixr2` fits, such as a fit that errored).

```{r boundary}
# The Emax model pushed ec50 to its lower boundary
isBoundaryFit(fitEmaxBoundaryIssue)

# The step-change model did not have a boundary issue
isBoundaryFit(fitStep)
```

# Choosing the best model by AIC

`getMinAICFit()` returns the fit with the lowest AIC. By default
(`excludeBoundary = TRUE`) it removes any boundary fits before comparing, and it
silently ignores any argument that cannot produce an AIC (for example, a model
that failed to estimate). You can pass fits as individual arguments or as a
list, and it emits a message when a boundary fit is removed.

```{r bestFit-choose}
bestFit <- getMinAICFit(fitEmaxBoundaryIssue, fitStep, fitLinear)
```

If every candidate is excluded or has no AIC, `getMinAICFit()` returns `NULL`
with a warning, so it is safe to call inside a larger report-building pipeline.

# Preparing for the report

## Put the models in a named list

By putting the models in a named list, the functions below can build more parts
of the report. The names become the model descriptions in the summary table.

```{r allFit-setup}
allFits <-
  list(
    "Emax model with additive residual error" = fitEmaxBoundaryIssue,
    "Step-change model with additive residual error" = fitStep,
    "Linear model with additive residual error" = fitLinear
  )
```

## Find your best model by AIC

```{r bestFit-list}
bestFit <- getMinAICFit(allFits)
```

## Summarize the best model with its equations and parameters

```{r bestFit-eqn}
knit_print(bestFit, inline = FALSE)
```

```{r bestFit-param}
pander::pander(bestFit$parFixed, caption = "Model parameters for the best-fit model")
```

## Summarize all models tested

`listModelsTested()` returns a data.frame with the model descriptions, their
AIC, and the change from the minimum AIC (dAIC). Models with a boundary issue
are flagged in an `Exclude` column and are left out of the dAIC calculation, and
the returned data.frame carries a `caption` attribute for pretty printing with
`pander::pander()`.

```{r list-models}
pander::pander(
  listModelsTested(allFits, caption = "Listing of all models tested.")
)
```
