---
title: "Nonparametric Bootstrap with runBootstrap()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nonparametric Bootstrap with runBootstrap()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

The nonparametric bootstrap is the most widely used method for quantifying parameter
uncertainty in population PK/PD modelling.  The idea is straightforward: resample
subjects with replacement from the original dataset, refit the model to each resampled
dataset, and take the empirical distribution of the resulting parameter estimates as an
approximation to the sampling distribution.  Unlike the asymptotic covariance matrix,
bootstrap confidence intervals do not assume that the likelihood surface is quadratic
near the maximum, and they naturally capture non-normality and boundary behaviour.

`runBootstrap()` automates this workflow for any `nlmixr2` fit:

1. **Resample** subjects with replacement (optionally stratified).
2. **Refit** the model to each bootstrap dataset using the same estimator as the
   original fit.
3. **Aggregate** parameter estimates across successful replicates and compute
   empirical confidence intervals.
4. **Attach** the results to the original fit object and write output files.

Partial runs are cached to disk so that an interrupted bootstrap can be resumed
without discarding completed replicates.

---

## Example: theophylline single-dose PK

We use `nlmixr2data::theo_sd` — 12 subjects after a single oral dose of theophylline
— with a one-compartment model that includes IIV on absorption, clearance, and volume.

```{r packages, message = FALSE}
library(nlmixr2)
library(nlmixr2data)
library(nlmixr2extra)
```

### 1. Fit the base model

```{r fit, message = FALSE, results = "hide"}
one.cmt <- function() {
  ini({
    tka    <- 0.45;  label("Log Ka")
    tcl    <- 1.00;  label("Log Cl")
    tv     <- 3.45;  label("Log V")
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v  ~ 0.1
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv  + eta.v)
    linCmt() ~ add(add.sd)
  })
}

fit <- nlmixr2(
  one.cmt,
  data    = nlmixr2data::theo_sd,
  est     = "focei",
  control = list(print = 0L)
)
```

```{r fit-summary}
fit$parFixedDf
```

---

### 2. Run the bootstrap

The most important arguments are:

| Argument | Default | Meaning |
|---|---|---|
| `nboot` | `200` | Number of bootstrap replicates |
| `nSampIndiv` | N subjects | Subjects per bootstrap dataset |
| `stratVar` | (none) | Variable to stratify on |
| `stdErrType` | `"perc"` | CI method: `"perc"` (percentile), `"sd"`, or `"se"` |
| `ci` | `0.95` | Confidence level |
| `restart` | `FALSE` | `TRUE` = discard cache, start fresh |
| `outputDir` | auto | Directory for cache and output files |
| `updateFit` | `TRUE` | Attach results to `fit$bootstrap` |
| `workers` | `NULL` | Sequential; set > 1 for parallel |

The example below uses `nboot = 5` for speed.  In practice, 200–1 000 replicates are
typical; 200 is usually sufficient for 95 % CI estimation in well-identified models.

```{r bootstrap, message = FALSE}
withr::with_tempdir({

  boot_result <- runBootstrap(
    fit,
    nboot      = 5,
    ci         = 0.95,
    restart    = TRUE,
    workers    = 1L
  )

})
```

`runBootstrap()` returns a named list:

| Element | Contents |
|---|---|
| `summary` | `parFixedDf` augmented with bootstrap columns |
| `results` | One row per successful replicate (also written to CSV) |
| `seed` | RNG state before resampling |
| `model` | Fit name used for the run |
| `outputDir` | Absolute path to the output directory |
| `timestamp` | When the function returned |

---

### 3. Read the summary

The `summary` element extends the original `parFixedDf` with six bootstrap columns:

```{r boot-summary}
boot_result$summary[, c(
  "Estimate", "Back-transformed",
  "Bootstrap CI Lower", "Bootstrap CI Upper",
  "Bootstrap %RSE"
)]
```

Key columns:

| Column | Meaning |
|---|---|
| `Bootstrap Estimate` | Mean of the bootstrap parameter estimates (model scale) |
| `Bootstrap SE` | SD of the bootstrap estimates (model scale) |
| `Bootstrap %RSE` | 100 × SE / \|mean estimate\| |
| `Bootstrap CI Lower/Upper` | Empirical percentile CI (back-transformed scale) |
| `Bootstrap Back-transformed` | Mean estimate on the back-transformed scale |

---

### 4. Accessing results on the fit object

When `updateFit = TRUE` (the default), `runBootstrap()` attaches a `$bootstrap` list
to the fit environment.  This is available after the function returns and persists for
the life of the `fit` object in your session:

```{r fit-bootstrap, eval = FALSE}
# Bootstrap summary table
fit$bootstrap$summary

# Omega CI matrix
fit$bootstrap$omegaSummary

# Empirical covariance matrix (thetas + omega lower triangle)
fit$bootstrap$covMatrix

# Correlation matrix
fit$bootstrap$corMatrix

# Bias: |original estimate - bootstrap mean|
fit$bootstrap$bias$parFixed
fit$bootstrap$bias$omega

# Metadata
fit$bootstrap$nboot
fit$bootstrap$ci
fit$bootstrap$outputDir
```

The bootstrap covariance matrix stored in `fit$bootstrap$covMatrix` can be activated
as the fit's working covariance matrix (used in simulations with parameter uncertainty)
via `nlmixr2est::setCov()`:

```{r set-cov, eval = FALSE}
# Switch to bootstrap-derived covariance
nlmixr2est::setCov(fit, "boot5")   # "bootN" where N = nboot

# Revert to the original FOCE-I covariance
nlmixr2est::setCov(fit, "linFim")
```

---

### 5. Per-replicate results

The `results` element (also written as `bootstrap_results.csv`) contains one row per
successful replicate with the model-scale estimate and back-transformed value for every
fixed-effect parameter:

```{r per-replicate}
head(boot_result$results)
```

This table is useful for computing custom summaries, visualising the bootstrap
distribution, or identifying outlier replicates.

---

### 6. Stratified bootstrap

When subjects fall into distinct groups with different sampling designs (e.g., sparse
vs. rich sampling, different study arms), it is important to preserve the proportion of
each group across bootstrap datasets.  Use `stratVar` to name the grouping column:

```{r stratified, eval = FALSE}
# Add a sparse/rich sampling flag to the data
theo_strat <- nlmixr2data::theo_sd
theo_strat$SPARSE <- ifelse(theo_strat$ID <= 6, "sparse", "rich")

fit_strat <- nlmixr2(one.cmt, data = theo_strat, est = "focei",
                     control = list(print = 0L))

withr::with_tempdir({
  runBootstrap(fit_strat, nboot = 200, stratVar = "SPARSE")
})
```

With stratification, the proportion of sparse and rich subjects in each bootstrap
dataset mirrors the original dataset exactly.

---

### 7. Resuming an interrupted run

`runBootstrap()` caches each completed replicate to disk.  If the run is interrupted,
re-call the function with the same `outputDir` and `restart = FALSE` (the default) to
pick up from where it stopped:

```{r resume, eval = FALSE}
# First partial run (e.g., interrupted after 50 of 200)
runBootstrap(fit, nboot = 200, restart = TRUE)

# Resume — only the remaining replicates are fitted
runBootstrap(fit, nboot = 200, restart = FALSE)
```

To start from scratch and discard any cached replicates, set `restart = TRUE`.

---

### 8. Increasing the replicate count

To add more replicates to an already completed run, simply increase `nboot` and
re-call with `restart = FALSE`.  The function detects the existing cache and only fits
the additional replicates:

```{r add-replicates, eval = FALSE}
# Start with 100
runBootstrap(fit, nboot = 100, restart = TRUE)

# Extend to 200 without redoing the first 100
runBootstrap(fit, nboot = 200, restart = FALSE)
```

---

### 9. Parallel execution

Each bootstrap replicate is an independent model fit, so parallelisation is
straightforward and can cut wall-clock time near-linearly up to the number of cores.
Set `workers` to the desired number of parallel processes:

```{r parallel, eval = FALSE}
withr::with_tempdir({
  runBootstrap(
    fit,
    nboot   = 200,
    workers = 4L,
    restart = TRUE
  )
})
```

`workers = NULL` (the default) uses the current `future` plan unchanged.
`workers = 1` forces sequential execution without requiring `future`.

---

### 10. Bootstrap diagnostic plot

`bootplot()` compares the empirical distribution of bootstrap ΔOFV values to a
chi-squared reference distribution, providing a graphical check of whether the
bootstrap sample size is adequate for the model complexity (Niebecker & Karlsson 2013).
The ΔOFV curve should lie on or below the chi-squared reference curve; a curve that
consistently exceeds the reference suggests the bootstrap may be underpowered.

```{r bootplot, eval = FALSE}
bootplot(fit)
```

---

## Output files

`runBootstrap()` writes the following files to the output directory:

| File | Contents |
|---|---|
| `boot_data_{N}.rds` | Resampled dataset for replicate N |
| `modelsEnsemble_{N}.rds` | Abbreviated fit summary for replicate N |
| `fitEnsemble_{N}.rds` | Full fit object for replicate N |
| `bootstrap_seed.rds` | Full RNG state before resampling |
| `bootstrap_results.csv` | Per-replicate parameter table (one row per replicate) |
| `bootstrap_summary.txt` | Human-readable summary with CIs |

The numbered output directory is named `<fitName>_boot_<N>` and is created
automatically in the working directory unless `outputDir` is specified explicitly.
Repeated fresh runs produce `fit_boot_1`, `fit_boot_2`, and so on; resuming reuses
the most recently numbered directory.

---

## Choosing `nboot` and `stdErrType`

**How many replicates?**  200 is a common minimum for 95 % CI estimation; 1 000 is
safer for narrow models or when tail behaviour matters.  A practical check: re-run
with twice as many replicates and verify that the CIs are stable.

**Which `stdErrType`?**

* `"perc"` (default) — empirical percentile method.  Requires no distributional
  assumption; recommended for most applications.
* `"sd"` — normal approximation using the bootstrap SD.  Appropriate when the
  parameter distribution is approximately normal.
* `"se"` — normal approximation using SE scaled by `nSampIndiv`.  Rarely needed.

For parameters that are bounded or skewed (variance components, steep absorption rate
constants), `"perc"` is strongly preferred.

---

## References

Efron, B. & Tibshirani, R.J. (1993). *An Introduction to the Bootstrap*. Chapman &
Hall.

Niebecker, R. & Karlsson, M.O. (2013). Are datasets for NLME models large enough for
a bootstrap to provide reliable parameter uncertainty distributions? *PAGE 22*,
Abstract 2899.

Wang, W., Hallow, K.M., & James, D.A. (2015). A tutorial on RxODE: simulating
differential equation pharmacometric models in R. *CPT: Pharmacometrics & Systems
Pharmacology*, 5, 3–10.
