---
title: "Log-Likelihood Profiling with runLLP()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Log-Likelihood Profiling with runLLP()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Introduction

Wald confidence intervals — computed as estimate ± 1.96 × SE — rest on the assumption
that the likelihood surface is quadratic near its maximum.  For non-linear mixed-effects
models this assumption is often violated, especially for parameters that are near a
boundary or whose profile is asymmetric (e.g., variance components, steep absorption
rate constants).

Log-likelihood profiling (Beal 1984; PsN `llp`) avoids this by tracing the actual
likelihood surface.  For each parameter in turn, the remaining parameters are
re-estimated at a grid of fixed values and the change in OFV (ΔOFV) is recorded.  The
95 % confidence bound is the parameter value at which

$$\Delta\text{OFV} = \chi^2_{0.95,1} \approx 3.84$$

`runLLP()` implements this algorithm for `nlmixr2` FOCE-I fits.  It supports:

* structural fixed effects (thetas) and residual-error thetas (e.g. `add.sd`)
* omega diagonal variance parameters (via `which = "eta.X"`)
* parallel execution via the `future` / `future.apply` packages

---

## Example: theophylline single-dose PK

We use the built-in `nlmixr2data::theo_sd` dataset — 12 subjects after a single oral
dose of theophylline.  The model is a one-compartment first-order absorption PK model
with IIV on the absorption rate constant and additive residual error.

```{r packages, message = FALSE}
library(nlmixr2)
library(nlmixr2data)
library(nlmixr2extra)
```

### 1. Fit the base model

```{r fit, message = FALSE, results = "hide"}
one.compartment <- function() {
  ini({
    tka    <- log(1.57)        # log absorption rate constant
    tcl    <- log(2.72)        # log clearance
    tv     <- fixed(log(31.5)) # log volume (fixed)
    eta.ka ~ 0.6               # IIV on absorption
    add.sd <- 0.7              # additive residual error SD
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl)
    v  <- exp(tv)
    cp <- linCmt()
    cp ~ add(add.sd)
  })
}

fit <- nlmixr2(
  one.compartment,
  data    = nlmixr2data::theo_sd,
  est     = "focei",
  control = list(print = 0L)
)
```

```{r fit-summary}
fit$parFixedDf
```

---

### 2. Profile structural thetas and residual error

`runLLP()` accepts a character vector of parameter names via `which`.  Here we profile
the two structural thetas (`tka`, `tcl`) and the residual-error parameter (`add.sd`).

`llpControl()` lets you tune the search.  The defaults are suitable for most
situations; the key options are:

| Argument | Default | Meaning |
|---|---|---|
| `ofvIncrease` | `qchisq(0.95, 1)` ≈ 3.84 | Target ΔOFV for 95 % CI |
| `normq` | 1.96 | Initial guess width = `normq × SE` |
| `rseTheta` | 30 | Fallback RSE (%) when SE is unavailable |
| `itermax` | 10 | Maximum iterations per bound |
| `ofvtol` | 0.005 | Convergence tolerance on ΔOFV |
| `workers` | `NULL` | Sequential; set > 1 for parallel |

```{r profile-theta}
ctrl <- llpControl(
  normq    = 1.96,
  rseTheta = 30,
  itermax  = 10,
  ofvtol   = 0.005
)

prof_theta <- runLLP(fit, which = c("tka", "tcl", "add.sd"), control = ctrl)
```

Printing the result shows the 95 % CI bounds, the MLE, and an asymmetry ratio (upper
half-width / lower half-width):

```{r print-theta}
print(prof_theta)
```

Extract the bounds as a matrix with `confint()`:

```{r confint-theta}
confint(prof_theta)
```

---

### 3. Visualise the profiles

`plot()` returns a `ggplot2` object with one panel per parameter.  The dashed
horizontal line marks the ΔOFV target; the vertical lines mark the MLE and CI bounds.

```{r plot-theta, fig.cap = "Likelihood profiles for tka, tcl, and add.sd"}
plot(prof_theta)
```

---

### 4. Profile an omega diagonal

Variance components often have highly asymmetric likelihood profiles because they are
bounded below by zero.  Pass the ETA name (e.g. `"eta.ka"`) to `which` — `runLLP()`
recognises omega diagonal parameters automatically and uses the appropriate
re-estimation mechanism.

```{r profile-omega}
prof_omega <- runLLP(fit, which = "eta.ka", control = ctrl)
print(prof_omega)
```

```{r confint-omega}
ci_omega <- confint(prof_omega)
ci_omega
```

Note that the lower bound is positive (variances cannot be negative) and the interval
is noticeably asymmetric — Wald intervals would under-cover in this region.

```{r plot-omega, fig.cap = "Likelihood profile for the eta.ka variance"}
plot(prof_omega)
```

---

### 5. Combined theta and omega profile

All parameters can be profiled in a single `runLLP()` call by listing both theta and
omega names together.  The return value is a single `nlmixr2Profile` object covering
all parameters.

```{r profile-all}
prof_all <- runLLP(
  fit,
  which   = c("tka", "tcl", "add.sd", "eta.ka"),
  control = ctrl
)

confint(prof_all)
```

```{r plot-all, fig.cap = "Combined profile for all four parameters", fig.height = 5}
plot(prof_all)
```

---

### 6. Parallel profiling

When many parameters need to be profiled, execution time can be reduced by running
each parameter in a separate worker.  Install the optional `future` and `future.apply`
packages and set `workers` in `llpControl()`:

```{r parallel, eval = FALSE}
# install.packages(c("future", "future.apply"))  # if not already installed

prof_parallel <- runLLP(
  fit,
  which   = c("tka", "tcl", "add.sd", "eta.ka"),
  control = llpControl(workers = 4L)
)
```

Sequential profiling (`workers = NULL` or `workers = 1`) does not require `future` or
`future.apply`.

---

## Working with the output

`prof_all` is an `nlmixr2Profile` object that inherits from `data.frame`.  Each row
corresponds to one model evaluation; the `Parameter` column identifies which parameter
was being profiled, `OFV` is the objective function value, and `profileBound` (when
non-`NA`) records the estimated CI bound for that row.

```{r output-structure}
head(prof_all[!is.na(prof_all$profileBound), c("Parameter", "OFV", "tka", "tcl", "add.sd", "eta.ka", "profileBound")])
```

Additional diagnostic information is stored as attributes:

```{r attributes}
# MLE values used as the reference point
attr(prof_all, "mle")

# Were any bounds near a parameter boundary?
attr(prof_all, "nearBound")

# Did any search hit the iteration limit?
attr(prof_all, "hitMaxIter")

# Upper / lower half-width ratio (> 1.3 triggers a warning)
attr(prof_all, "intervalRatio")
```

---

## References

Beal, S.L. (1984). Population pharmacokinetic data and parameter estimation based on
their first two statistical moments. *Drug Metabolism Reviews*, 15, 173–193.

Lindbom, L., Ribbing, J., & Jonsson, E.N. (2004). Perl-speaks-NONMEM (PsN) — a Perl
module for NONMEM related programming. *Computer Methods and Programs in Biomedicine*,
75, 85–94.
