---
title: "Sampling Importance Resampling with runSIR()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sampling Importance Resampling with runSIR()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

Standard uncertainty estimates for non-linear mixed-effects model parameters rely on the
covariance matrix produced by the covariance step (an approximation to the inverse
Fisher information matrix).  This approximation performs well when the likelihood
surface is nearly quadratic near the maximum — a condition that often fails for
variance components, highly correlated parameters, or models with a restricted
parameter space.

Sampling Importance Resampling (SIR; Dosne et al. 2013) is a simulation-based
alternative that samples directly from the parameter uncertainty distribution.  The
algorithm:

1. **Samples** parameter vectors from a proposal distribution centred on the MLE and
   using the asymptotic covariance as its shape.
2. **Evaluates** each sampled vector by re-running the model with those parameters
   fixed (one OFV evaluation per sample, equivalent to NONMEM `MAXEVAL=0`).
3. **Weights** each sample by the ratio of the target density (the likelihood) to the
   proposal density (importance weighting).
4. **Resamples** without replacement according to those weights.
5. **Updates** the proposal covariance from the empirical covariance of the resampled
   vectors and repeats for the next iteration.

After convergence the resampled vectors represent draws from the parameter uncertainty
distribution, and empirical percentiles give non-parametric confidence intervals that
require no asymptotic assumption.

`runSIR()` implements this algorithm for `nlmixr2` FOCE-I fits, following the PsN
`sir` workflow where practical.

---

## Prerequisites

`runSIR()` requires a fit with a successful covariance step — the initial proposal is
derived from `fit$cov`.  If your fit was run without a covariance step, re-run with
`covMethod = "r"` (or another appropriate method) before calling `runSIR()`.

SIR also requires at least two random effects (ETAs) so that a multivariate proposal
can be constructed for the omega block.

```{r packages, message = FALSE}
library(nlmixr2)
library(nlmixr2data)
library(nlmixr2extra)
```

---

## Example: theophylline single-dose PK

We use `nlmixr2data::theo_sd` — 12 subjects after a single oral dose — with a
one-compartment model that includes IIV on absorption, clearance, and volume.

### 1. Fit the base model

The covariance step runs automatically with FOCE-I.  Request NPDE and CWRES residuals
via `table` for later diagnostics if needed.

```{r fit, message = FALSE, results = "hide"}
three_eta_one_cmt <- function() {
  ini({
    tka    <- 0.45    # log absorption rate constant (1/h)
    tcl    <- 1.00    # log clearance (L/h)
    tv     <- 3.45    # log central volume (L)
    eta.ka ~ 0.6      # IIV on absorption
    eta.cl ~ 0.3      # IIV on clearance
    eta.v  ~ 0.1      # IIV on volume
    add.sd <- 0.7     # additive residual error (mg/L)
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv  + eta.v)
    linCmt() ~ add(add.sd)
  })
}

fit <- nlmixr2(
  three_eta_one_cmt,
  data    = nlmixr2data::theo_sd,
  est     = "focei",
  control = list(print = 0L),
  table   = list(npde = TRUE, cwres = TRUE)
)
```

```{r fit-summary}
fit$parFixedDf
round(as.matrix(fit$omega), 4)
```

---

### 2. Run SIR

The most important arguments are:

| Argument | Default | Meaning |
|---|---|---|
| `nSamples` | `c(1000, 1000, 1000, 2000, 2000)` | Samples per iteration |
| `nResample` | `c(200, 400, 500, 1000, 1000)` | Resamples per iteration |
| `thetaInflation` | `1` | Variance multiplier for theta proposal |
| `omegaInflation` | `1` | Variance multiplier for omega proposal |
| `capCorrelation` | `0.8` | Cap on absolute pairwise correlations |
| `capResampling` | `1` | Max times one sample may be redrawn |
| `recenter` | `TRUE` | Shift proposal mean to best sample when dOFV < 0 |
| `boxcox` | `TRUE` | Use Box-Cox transformed empirical covariance |
| `workers` | `NULL` | Sequential; set > 1 for parallel OFV evaluation |
| `directory` | `NULL` | Output folder (auto-named if `NULL`) |
| `recover` | `TRUE` | Resume from saved state if found |

The example below uses a two-iteration schedule with small sample counts for
illustration.  In a real analysis, use the PsN-recommended defaults (1 000 / 2 000
samples, 200 / 1 000 resamples) or increase counts until the resampled covariance
stabilises.

```{r sir, message = FALSE}
set.seed(20260420)

sir_result <- runSIR(
  fit,
  nSamples       = c(16L, 16L),
  nResample      = c(8L,  8L),
  thetaInflation = 1,
  omegaInflation = 1,
  sigmaInflation = 1,
  capCorrelation = 0.8,
  capResampling  = 1,
  recenter       = TRUE,
  boxcox         = TRUE,
  workers        = 1L,
  directory      = file.path(
    tempdir(), paste0("sir_vignette_", format(Sys.time(), "%Y%m%d_%H%M%S"))
  ),
  recover        = FALSE
)
```

---

### 3. Inspect the results

`print()` shows the final parameter uncertainty table and per-iteration diagnostics:

```{r print-sir}
print(sir_result)
```

Key columns in the summary:

| Column | Meaning |
|---|---|
| `estimate` | MLE from the original fit |
| `sd` | Empirical SD of the resampled vectors |
| `rse` | 100 × sd / \|estimate\| |
| `p2.5`, `p97.5` | 2.5th and 97.5th empirical percentiles (95 % CI) |
| `p50` | Median of the resampled distribution |

The iteration diagnostics report how many samples were drawn, how many OFV evaluations
succeeded, and how many vectors were resampled.  A healthy run has few failures
(`nFailed` ≈ 0) and no negative dOFV values (`nNegativeDOFV` = 0; if present, the
proposal mean is recentered automatically when `recenter = TRUE`).

---

### 4. Re-compute the summary from the resampled matrix

`sirSummary()` recomputes the table from the raw resampled matrix, which is also
useful when you want to summarise a subset of iterations or apply the function to a
matrix from an external source.

```{r sir-summary}
final_mat  <- attr(sir_result, "resampledMat")
sir_summ   <- sirSummary(final_mat, fit)
print(sir_summ)
```

Empirical covariance and correlation matrices between parameters are available as
attributes:

```{r cov-cor}
round(attr(sir_summ, "covMatrix"), 6)
round(attr(sir_summ, "corMatrix"), 4)
```

---

### 5. Diagnostic plots

`plot()` returns a `ggplot2` object.  Three views are available:

#### Parameter distributions

Histograms of the final resampled vectors with the MLE marked in orange.  The shape
reveals whether the posterior is symmetric and whether any parameter is near a bound.

```{r plot-params, fig.cap = "Final resampled parameter distributions"}
plot(sir_result, type = "parameters", bins = 12L)
```

#### dOFV distributions

Histograms of ΔOFV = OFV(sample) − OFV(fit) per iteration.  A good proposal yields a
distribution concentrated near 0–10; a very flat or right-skewed dOFV histogram
suggests the proposal is too narrow or too wide.

```{r plot-dofv, fig.cap = "dOFV distributions by iteration"}
plot(sir_result, type = "dofv", bins = 12L)
```

#### Resampling probabilities

Bar chart of resampling probabilities per sample per iteration.  A healthy run has
many samples with non-negligible probability; a run dominated by a few high-weight
samples indicates a poor proposal (too wide).

```{r plot-resample, fig.cap = "Resampling probabilities by iteration"}
plot(sir_result, type = "resampling")
```

---

### 6. Output files

`runSIR()` writes several files to the output directory:

```{r list-files}
list.files(attr(sir_result, "outputDir"))
```

| File | Contents |
|---|---|
| `raw_results_sir_iteration{N}.csv` | Full sample/weight/resample table for each iteration |
| `summary_iterations.csv` | Per-iteration diagnostic summary |
| `sample_rejection_summary.txt` | Counts of rejected samples (bounds, omega non-PD) |
| `sir_results.csv` | Final parameter summary (same as the printed table) |
| `sir_state.rds` | Full internal state for recovery or adding iterations |

---

### 7. Advanced use

#### Resuming an interrupted run

By default `recover = TRUE` so that if a long SIR run is interrupted, re-calling
`runSIR()` with the same `directory` will pick up from the last completed iteration:

```{r recover, eval = FALSE}
sir_result <- runSIR(fit, directory = "my_sir_run", recover = TRUE)
```

#### Adding iterations

To append more iterations to a completed run, use `addIterations = TRUE` with a new
sample schedule:

```{r add-iter, eval = FALSE}
sir_result <- runSIR(
  fit,
  nSamples      = c(2000L, 2000L),
  nResample     = c(1000L, 1000L),
  directory     = "my_sir_run",
  addIterations = TRUE
)
```

#### Parallel OFV evaluation

OFV evaluations are the bottleneck.  Setting `workers > 1` runs them in parallel using
the `future` / `future.apply` packages:

```{r parallel, eval = FALSE}
# install.packages(c("future", "future.apply"))  # if not already installed

sir_result <- runSIR(
  fit,
  nSamples  = c(1000L, 1000L, 1000L, 2000L, 2000L),
  nResample = c(200L,  400L,  500L,  1000L, 1000L),
  workers   = 4L
)
```

Sequential evaluation (`workers = NULL` or `workers = 1`) does not require `future` or
`future.apply`.

---

## Choosing the schedule

The PsN documentation recommends the default five-iteration schedule
`nSamples = c(1000, 1000, 1000, 2000, 2000)` /
`nResample = c(200, 400, 500, 1000, 1000)`.  The rule of thumb is that each
iteration's sample-to-resample ratio should be approximately 5 : 1.  The final
resampled covariance should be stable across iterations — if it is still changing
noticeably between the last two iterations, add more iterations or increase sample
counts.

For parameters that are far from normal (e.g. highly constrained variance components),
setting `thetaInflation` or `omegaInflation` to 1.5–2 widens the initial proposal and
reduces the risk of importance-weight collapse.

---

## References

Dosne, A.-G., Bergstrand, M., & Karlsson, M.O. (2013). An automated sampling
importance resampling procedure for estimating parameter uncertainty. *PAGE 22*,
Abstract 2907.

Dosne, A.-G., Bergstrand, M., Harling, K., & Karlsson, M.O. (2016). Improving the
estimation of parameter uncertainty distributions in nonlinear mixed effects models
using sampling importance resampling. *Journal of Pharmacokinetics and
Pharmacodynamics*, 43(6), 583–596.

Lindbom, L., Ribbing, J., & Jonsson, E.N. (2004). Perl-speaks-NONMEM (PsN) — a Perl
module for NONMEM related programming. *Computer Methods and Programs in Biomedicine*,
75, 85–94.
