---
title: "Stepwise Covariate Modelling with runSCM()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stepwise Covariate Modelling with runSCM()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Introduction

Stepwise Covariate Modelling (SCM) is the standard automated covariate selection
procedure in population PK/PD analysis.  It uses a likelihood-ratio test to evaluate
whether adding a covariate-parameter relationship significantly improves model fit,
and proceeds in two phases:

1. **Forward inclusion**: starting from the base model, the candidate relationship
   that produces the largest statistically significant ΔOFV (OFV decrease) is added
   at each step.  The process repeats until no remaining candidate meets the threshold.
2. **Backward elimination**: starting from the forward-final model, each included
   relationship is tested for removal.  A relationship is dropped if its removal
   does not produce a statistically significant ΔOFV increase (at a stricter
   threshold than forward).

`runSCM()` implements both phases for `nlmixr2` fits.  It automatically generates
the covariate expressions in the model body — centering continuous covariates at their
observed median, creating indicator columns for categorical covariates — so no data
pre-processing is required.

---

## Example: warfarin PK covariate search

We use `nlmixr2data::warfarin`, filtered to PK observations only (`dvid == "cp"`), with
covariates body weight (`wt`, continuous) and sex (`sex`, categorical).

```{r packages, message = FALSE}
library(nlmixr2data)
library(nlmixr2extra)
```

### 1. Prepare the data

Pass the **original, untransformed** dataset to `runSCM()`.  All covariate centering
and other shape transformations are generated inside the model body by the SCM
machinery; applying them to the data first would lead to double-transformation.

```{r data}
pkdata <- warfarin[warfarin$dvid == "cp", ]

d_subj <- pkdata[!duplicated(pkdata$id), ]
cat(sprintf("Subjects: %d   Rows: %d\n", length(unique(pkdata$id)), nrow(pkdata)))
cat(sprintf("wt  median: %.1f  range: %.1f – %.1f\n",
            median(d_subj$wt), min(d_subj$wt), max(d_subj$wt)))
cat(sprintf("sex levels: %s\n",
            paste(names(table(d_subj$sex)), table(d_subj$sex), sep = "=", collapse = ", ")))
```

### 2. Fit the base model

The base model contains no covariate terms — `runSCM()` adds them.

```{r fit, message = FALSE, results = "hide"}
warf_pk <- function() {
  ini({
    tka     <- log(1.15)   # log absorption rate constant (h^-1)
    tcl     <- log(0.135)  # log clearance (L/h)
    tv      <- log(7.0)    # log volume of distribution (L)
    eta.ka  ~ 0.40
    eta.cl  ~ 0.25
    eta.v   ~ 0.10
    prop.err <- 0.10
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv  + eta.v)
    linCmt() ~ prop(prop.err)
  })
}

fit_base <- nlmixr2(
  warf_pk, pkdata,
  est     = "saem",
  control = nlmixr2est::saemControl(print = 0, nBurn = 200, nEm = 100),
  table   = nlmixr2est::tableControl(cwres = TRUE)
)
```

```{r fit-summary}
fit_base$parFixedDf
cat(sprintf("Base OFV: %.3f\n", fit_base$objf))
```

---

### 3. Run the SCM

#### Specifying candidates with `pairsVec`

The most explicit way to define the search space is `pairsVec`: a list of
`list(var=, covar=, shapes=)` items.  Each item names the PK parameter (`var`),
the covariate (`covar`), and optionally the functional shapes to test.  When
`shapes` is omitted the global `shapes` argument applies (default `"power"`).

Categorical covariates are listed in `catvarsVec`; `runSCM()` automatically
creates binary indicator columns and tests each non-reference level.

The `data` argument must be supplied explicitly whenever the base model does not
reference a covariate column — otherwise `nlme::getData(fit)` may drop it.

```{r scm-pairs}
pairs_list <- list(
  list(var = "v",  covar = "wt",  shapes = "power"),
  list(var = "v",  covar = "sex"),
  list(var = "cl", covar = "wt",  shapes = "power"),
  list(var = "cl", covar = "sex")
)
```

```{r scm, message = FALSE, results = "hide"}
scm <- runSCM(
  fit        = fit_base,
  data       = pkdata,
  pairsVec   = pairs_list,
  catvarsVec = "sex",
  pVal       = list(fwd = 0.05, bck = 0.01),
  searchType = "scm",
  saveModels = TRUE,
  verbose    = FALSE,
  restart    = TRUE,
  print      = 0,
  workers    = 1L,
  confirm    = FALSE
)
```

#### `runSCM()` argument reference

| Argument | Default | Meaning |
|---|---|---|
| `varsVec` | `NULL` | PK parameters to test (used with `covarsVec`; ignored when `pairsVec` supplied) |
| `covarsVec` | `NULL` | Continuous covariates (ignored when `pairsVec` supplied) |
| `catvarsVec` | `NULL` | Categorical covariates; indicator columns created automatically |
| `pairsVec` | `NULL` | Explicit parameter-covariate pairs; overrides `varsVec`/`covarsVec` |
| `shapes` | `"power"` | Shape(s) for continuous covariates: `"power"`, `"lin"`, `"log"`, `"identity"` |
| `pVal` | `list(fwd=0.05, bck=0.01)` | Forward and backward p-value thresholds |
| `searchType` | `"scm"` | `"scm"` (both phases), `"forward"`, or `"backward"` |
| `data` | `NULL` | Dataset; required when base model omits covariate columns |
| `inits` | `list()` | Initial theta estimates for covariate parameters, by shape |
| `missingToken` | `NA` | Sentinel value for missing covariate observations |
| `catCutoff` | `0.05` | Minimum non-reference level prevalence to test |
| `includedRelations` | `NULL` | Relations forced into the backward-start model |
| `saveModels` | `TRUE` | Write fitted models to `outputDir` |
| `verbose` | `FALSE` | Print full candidate tables at each step |
| `restart` | `FALSE` | Discard existing cache and start fresh |
| `workers` | `NULL` | Sequential; set > 1 for parallel candidate fitting |
| `confirm` | `TRUE` | Prompt for confirmation in interactive sessions |

---

### 4. Interpreting the results

`runSCM()` returns a named list with three elements:

```{r return-structure}
names(scm)
```

#### Summary table

`summaryTable` contains one row per candidate tested across all steps and both phases,
with the columns most useful for review:

```{r summary-table}
cols <- intersect(
  c("searchType", "step", "covar", "var", "shape",
    "deltObjf", "pchisqr", "included"),
  colnames(scm$summaryTable)
)
print(scm$summaryTable[, cols])
```

Key columns:

| Column | Meaning |
|---|---|
| `searchType` | `"forward"` or `"backward"` |
| `step` | Step number within the phase |
| `covar` | Covariate name |
| `var` | PK parameter |
| `shape` | Functional shape tested |
| `deltObjf` | ΔOFV (positive = improvement in forward; positive = deterioration in backward) |
| `pchisqr` | p-value from chi-squared test (1 df) |
| `bsvReduction` | Reduction in BSV for `var` (%) |
| `covarEffect` | Estimated effect magnitude at the covariate extremes |
| `included` | `"yes"` (accepted), `"no"` (rejected), `"dropped"` (removed in backward) |

#### Forward search detail

```{r fwd-detail}
fwd_tbl <- scm$resFwd[[2]]
if (!is.null(fwd_tbl) && nrow(fwd_tbl) > 0) {
  print(fwd_tbl[fwd_tbl$included == "yes",
                intersect(c("step", "covar", "var", "shape", "deltObjf",
                            "pchisqr", "bsvReduction"), colnames(fwd_tbl))])
} else {
  cat("No covariates accepted in forward search.\n")
}
```

#### Backward search detail

```{r bck-detail}
bck_tbl <- scm$resBck[[2]]
if (!is.null(bck_tbl) && nrow(bck_tbl) > 0) {
  print(bck_tbl[, intersect(c("step", "covar", "var", "shape", "deltObjf",
                               "pchisqr", "included"), colnames(bck_tbl))])
} else {
  cat("No backward elimination steps.\n")
}
```

#### Final model

The final model fit is in `resBck[[1]]` after a full SCM (or `resFwd[[1]]` after a
forward-only search):

```{r final-model}
fit_final <- scm$resBck[[1]]
fit_final$parFixedDf
cat(sprintf("Final OFV: %.3f   ΔOFV vs base: %.3f\n",
            fit_final$objf, fit_base$objf - fit_final$objf))
```

---

### 5. Alternative: specify all combinations with `varsVec` and `covarsVec`

When the search space is a full Cartesian product of parameters × covariates, use
`varsVec` and `covarsVec` instead of enumerating every pair:

```{r scm-auto, eval = FALSE}
scm_auto <- runSCM(
  fit        = fit_base,
  data       = pkdata,
  varsVec    = c("cl", "v"),
  covarsVec  = "wt",
  catvarsVec = "sex",
  shapes     = c("power", "lin"),
  pVal       = list(fwd = 0.05, bck = 0.01),
  confirm    = FALSE
)
```

This tests `wt~cl`, `wt~v`, `sex~cl`, and `sex~v`, each with power and linear shapes
for the continuous covariate — eight candidates in total.

---

### 6. Covariate shapes

`runSCM()` supports four built-in shapes for continuous covariates.  All use the
**observed median** as the centering value, computed automatically from the data.

| Shape | Expression in model body | Interpretation |
|---|---|---|
| `"power"` | `log(cov / median(cov))` | Allometric / power-law relationship |
| `"lin"` | `cov − median(cov)` | Linear deviation from the median |
| `"log"` | `log(cov)` | Log-transformed covariate (no centering) |
| `"identity"` | `cov` | Raw covariate (no transformation) |

Categorical covariates always use the `"cat"` shape: a binary indicator column
`cov_<level>` multiplied by a theta parameter.  The reference level is the most
frequent category.  Levels representing fewer than `catCutoff` (default 5 %) of
subjects are merged with the reference and not tested.

Custom shapes can be added via the `customShapes` argument — a named list of functions
`function(col, center, level)` that return a character expression string.

---

### 7. Forced inclusions

`includedRelations` forces specific covariate relationships into the model at the start
of backward elimination — even if they were not accepted during forward inclusion.
This mirrors PsN's `[included_relations]` block:

```{r included, eval = FALSE}
scm_inc <- runSCM(
  fit               = fit_base,
  data              = pkdata,
  pairsVec          = pairs_list,
  catvarsVec        = "sex",
  searchType        = "scm",
  includedRelations = list(list(var = "cl", covar = "wt", shapes = "power")),
  confirm           = FALSE
)
```

The `wt~cl` power relationship will be present at the start of backward elimination
regardless of whether it was significant in the forward phase.  It remains eligible
for removal during backward elimination like any other covariate.

---

### 8. Forward-only or backward-only search

Set `searchType` to run a single phase:

```{r single-phase, eval = FALSE}
# Forward only — stop after inclusion phase
scm_fwd <- runSCM(
  fit        = fit_base,
  data       = pkdata,
  pairsVec   = pairs_list,
  catvarsVec = "sex",
  searchType = "forward",
  confirm    = FALSE
)
fit_after_fwd <- scm_fwd$resFwd[[1]]

# Backward only — apply elimination to a model that already contains covariates
# (requires a fit that already has covariate terms)
scm_bck <- runSCM(
  fit        = fit_after_fwd,
  data       = pkdata,
  pairsVec   = pairs_list,
  catvarsVec = "sex",
  searchType = "backward",
  pVal       = list(fwd = 0.05, bck = 0.01),
  confirm    = FALSE
)
```

---

### 9. Parallel candidate fitting

Each candidate model at a given step is an independent fit, so parallelisation gives
a near-linear speed-up up to the number of candidates.  Install the optional
`future`, `future.apply`, and `progressr` packages and set `workers`:

```{r parallel, eval = FALSE}
# install.packages(c("future", "future.apply", "progressr"))

scm_par <- runSCM(
  fit        = fit_base,
  data       = pkdata,
  pairsVec   = pairs_list,
  catvarsVec = "sex",
  searchType = "scm",
  workers    = 4L,
  confirm    = FALSE
)
```

Sequential execution (`workers = NULL` or `workers = 1`) requires no additional
packages.

> **Note**: parallel SCM workers load from the **installed** package.  If you are
> developing with `devtools::load_all()`, install the package first with
> `devtools::install()` before using `workers > 1`.

---

### 10. Missing covariate values

When a covariate has missing observations, set `missingToken` to the sentinel used in
the data (e.g. `-99` or `"."`) in addition to `NA`.  `runSCM()` wraps the covariate
expression in an `ifelse()` guard that imputes the population-typical value:

* **Continuous**: imputes the median (log(median/median) = 0 for power shape; 0 for linear).
* **Categorical**: imputes the mode (most frequent level).

```{r missing, eval = FALSE}
# Data where missing wt is coded as -99
scm_miss <- runSCM(
  fit          = fit_base,
  data         = pkdata,
  pairsVec     = pairs_list,
  catvarsVec   = "sex",
  missingToken = -99,
  confirm      = FALSE
)
```

---

## Output files

When `saveModels = TRUE` (the default), `runSCM()` writes the following files to the
output directory (auto-named `<fitName>_scm_<N>` in the working directory):

| File | Contents |
|---|---|
| `scm_step_summary.csv` | Best candidate per step (one row per step) |
| `scm_all_candidates.csv` | Every model tested across all steps |
| `scm_log.txt` | Human-readable run log with header metadata |
| `fwd_step{N}_<rel>.rds` | Fitted model for each forward candidate |
| `bck_step{N}_<rel>.rds` | Fitted model for each backward candidate |

Set `saveModels = FALSE` to run without writing any files, for example when
exploring the search space in a scratch session.

---

## Tips for a robust SCM

**Sample size**: the SCM uses a chi-squared approximation with 1 degree of freedom.
The rule of thumb is at least 50–100 subjects for a reliable test; sparse data leads
to inflated type I error in forward inclusion.

**p-value thresholds**: the conventional PsN defaults are `fwd = 0.05` and
`bck = 0.01`.  In exploratory analyses, a more liberal forward threshold (e.g. 0.10)
avoids missing important relationships at the cost of higher false-positive rates.

**Shape choice**: the power shape (`log(cov/median)`) is appropriate for weight and
other allometric predictors.  Use `shapes = c("power", "lin")` to test both and let
the data decide.

**Warm-starting**: when a covariate is accepted, the estimated theta from that step
is automatically used as the starting estimate for all subsequent steps that involve
the same covariate shape, which speeds convergence and improves numerical stability.

**Categorical covariates**: pass `catvarsVec` rather than pre-creating dummy columns.
`runSCM()` handles level detection, reference-level selection, and indicator column
creation automatically.

---

## References

Jonsson, E.N. & Karlsson, M.O. (1998). Automated covariate model building within
NONMEM. *Pharmaceutical Research*, 15(9), 1463–1468.

Lindbom, L., Ribbing, J., & Jonsson, E.N. (2004). Perl-speaks-NONMEM (PsN) — a Perl
module for NONMEM related programming. *Computer Methods and Programs in Biomedicine*,
75, 85–94.

Ribbing, J. & Jonsson, E.N. (2004). Power, selection bias and predictive performance
of the population pharmacokinetic covariate model. *Journal of Pharmacokinetics and
Pharmacodynamics*, 31(2), 109–134.
