---
title: "Likelihood Profiling for NLME Models"
author: "Bill Denney"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Likelihood Profiling for NLME Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
library(nlmixr2save)
## A cached fit is restored by sourcing the R code nlmixr2save saved with it,
## and that code calls `lotri()` and `foceiControl()` unqualified -- so both
## packages must be attached before any `:=` loads from the cache.
library(lotri)
library(nlmixr2est)
## Cache each fit and each profile with the `:=` operator (nlmixr2save): they
## are stored under vignettes/cache/ with a per-article prefix and loaded as-is
## on the website build.  See vignettes/precompute.R.
options(nlmixr2save.dir = "../cache",
        nlmixr2save.prefix = "likelihood-profiling-",
        nlmixr2save.check = FALSE)
```

# Introduction

Wald-based confidence intervals (CIs), which are the default output from
`nlmixr2`, rely on the assumption that the parameter estimates are normally
distributed. For parameters near boundaries, or for variance components of
random effects, this assumption often breaks down and the resulting CIs can
be misleading.

**Likelihood profiling** provides a more reliable alternative. Rather than
approximating the likelihood surface with a quadratic, it evaluates the actual
objective function (OFV) at a grid of parameter values, holding each parameter
fixed in turn and re-estimating the rest. A parameter's confidence interval is
the set of values whose OFV increase relative to the minimum is below a
threshold -- by default $\chi^2_{0.95,1} \approx 3.84$.

`nlmixr2extra` provides two profiling methods via the `profile()` generic:

| Method | Function | Description |
|--------|----------|-------------|
| `"llp"` | `profileLlp()` | Adaptive log-likelihood profiling -- searches for the exact OFV boundary |
| `"fixed"` | `profileFixed()` | Evaluates the OFV at user-supplied fixed values |

# Quick start

## Define and fit a model

We use the single-dose theophylline dataset (`theo_sd`) from `nlmixr2data`.

```{r fit, message=FALSE, warning=FALSE}
library(nlmixr2extra)
library(ggplot2)

oneCmt <- function() {
  ini({
    tka  <- log(1.57)   # log absorption rate constant
    tcl  <- log(2.72)   # log clearance
    tv   <- log(31.5)   # log volume (fixed)
    eta.ka ~ 0.6
    eta.cl ~ 0.09
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv)
    cp <- linCmt()
    cp ~ add(add.sd)
  })
}

fit := nlmixr2(oneCmt, data = nlmixr2data::theo_sd,
               est = "focei", control = list(print = 0))
fit
```

## Profile all parameters

```{r profile-all}
# Profiles tka and tcl; eta.ka, eta.cl, and add.sd are also included.
profAll := profile(fit)
profAll
```

## Profile a single parameter

```{r profile-single}
profTka := profile(fit, which = "tka")
profTka
```

# The log-likelihood profiling method (`"llp"`)

`profileLlp()` uses an adaptive algorithm:

1. Start from the point estimate and step in both directions.
2. At each step, fix the target parameter and re-estimate all others.
3. Stop when the OFV change is within `ofvtol` of the desired `ofvIncrease`,
   or when the next parameter step would not change its rounded value to
   `paramDigits` significant figures.

## Control options

All LLP settings are set via `llpControl()`:

```{r llpcontrol}
# Default settings
ctrl <- llpControl(
  ofvIncrease   = qchisq(0.95, df = 1),  # ~3.84 for 95% CI
  rseTheta      = 30,    # starting step size as % of the estimate
  itermax       = 10,    # max iterations per direction
  ofvtol        = 0.005, # tolerance on OFV difference
  paramDigits   = 3      # significant digits for convergence
)

# a separate name: each `:=` target is its own cache entry
profTkaCtrl := profile(fit, which = "tka", control = ctrl)
```

To get a 90% CI instead of 95%, lower the threshold:

```{r llpcontrol-90}
profTka90 := profile(fit, which = "tka",
                     control = list(ofvIncrease = qchisq(0.90, df = 1)))
```

## Interpreting the output

The returned data frame has one row per model evaluation. Key columns:

- `Parameter` -- which parameter was fixed on this row
- `OFV` -- the objective function value at that step
- `profileBound` -- present on the boundary rows; its absolute value is the
  `ofvIncrease` threshold and its sign indicates lower (`-`) or upper (`+`)
- One column per model parameter showing its estimate at each step

```{r interpret}
# Show rows near the boundary
profTka[!is.na(profTka$profileBound), ]
```

The rows where `profileBound` is non-`NA` give the profile CI limits directly
from the `tka` column.

# Plotting the profile

The profile data frame contains everything needed for a profile plot. A
standard presentation shows OFV - OFV_min on the y-axis and the parameter
value on the x-axis, with a horizontal reference line at the CI threshold.

```{r plot}
# Compute Delta OFV relative to the minimum
ofvMin <- min(profTka$OFV, na.rm = TRUE)
profTka$dOFV <- profTka$OFV - ofvMin

# The two boundary rows (one per direction)
bounds <- profTka[!is.na(profTka$profileBound), ]

ggplot(profTka, aes(x = tka, y = dOFV)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = qchisq(0.95, 1), linetype = "dashed", colour = "red") +
  geom_vline(xintercept = bounds$tka, linetype = "dotted", colour = "blue") +
  labs(
    x     = "tka",
    y     = expression(Delta * "OFV"),
    title = "Likelihood profile for tka",
    subtitle = sprintf(
      "95%% profile CI: [%.3f, %.3f]",
      min(bounds$tka), max(bounds$tka)
    )
  ) +
  theme_bw()
```

A symmetric, approximately parabolic profile indicates the Wald CI is reliable.
An asymmetric or flat profile indicates non-normality; in these cases the
profile CI is more trustworthy.

# The fixed-point method (`"fixed"`)

`profileFixed()` evaluates the OFV at a user-supplied grid rather than
searching adaptively. It is useful when you want to reproduce a specific grid
(e.g. for a table or a plot comparing multiple models) or when the adaptive
algorithm struggles.

## Evaluating a uniform grid

```{r fixed}
# Evaluate tka at seven equally spaced values around the point estimate
tkaEst <- fit$theta["tka"]

grid <- data.frame(tka = seq(tkaEst - 0.6, tkaEst + 0.6, length.out = 7))

profGrid := profile(fit, which = grid, method = "fixed")
profGrid[, c("Parameter", "tka", "OFV")]
```

## Two-parameter joint profile

`profileFixed()` accepts a data frame with multiple columns. Each row fixes
all named parameters simultaneously, allowing you to trace joint profile
contours.

```{r fixed-joint}
grid2d <- expand.grid(
  tka = seq(tkaEst - 0.4, tkaEst + 0.4, length.out = 5),
  tcl = seq(nlmixr2est::fixef(fit)[["tcl"]] - 0.2,
            nlmixr2est::fixef(fit)[["tcl"]] + 0.2, length.out = 5)
)

profJoint := profile(fit, which = grid2d, method = "fixed")

ggplot(profJoint, aes(x = tka, y = tcl,
                       fill = OFV - min(OFV, na.rm = TRUE))) +
  geom_tile() +
  scale_fill_viridis_c(name = expression(Delta * "OFV")) +
  geom_contour(aes(z = OFV - min(OFV, na.rm = TRUE)),
               breaks = qchisq(0.95, 2), colour = "white") +
  labs(title = "Joint profile: tka x tcl",
       subtitle = "White contour = 95% joint confidence region") +
  theme_bw()
```

# Comparing profile CIs with Wald CIs

Profile CIs are wider than Wald CIs when the likelihood surface is asymmetric.
This is common for variance parameters and log-scale parameters near zero.

```{r compare}
# Wald CI from the fit
waldCi <- confint(fit)  # uses the covariance matrix

# Extract profile CI from the boundary rows
boundRows <- profAll[!is.na(profAll$profileBound), ]

# For each parameter, the two boundRows give lower and upper limits
profileCi <- tapply(
  seq_len(nrow(boundRows)),
  boundRows$Parameter,
  function(idx) {
    vals <- boundRows[idx, boundRows$Parameter[idx[1]]]
    c(lower = min(vals), upper = max(vals))
  }
)

# Show side-by-side
do.call(rbind, lapply(names(profileCi), function(p) {
  data.frame(
    parameter   = p,
    wald_lower  = waldCi[p, 1],
    wald_upper  = waldCi[p, 2],
    prof_lower  = profileCi[[p]]["lower"],
    prof_upper  = profileCi[[p]]["upper"]
  )
}))
```

# Tips

- **Start with `which`**: profiling all parameters is time-consuming. Profile
  only the parameters of primary interest first.
- **Increase `itermax`** if you see "aborted due to too many iterations". The
  default of 10 is conservative; 20-30 is often sufficient for variance
  parameters.
- **Adjust `rseTheta`**: if the initial step is too large (OFV jumps past the
  target immediately) reduce `rseTheta`. If it is too small (many steps needed
  before the boundary is reached) increase it.
- **Fixed-point fallback**: when `profileLlp()` fails to converge for a
  particular parameter, `profileFixed()` on a coarse grid around the estimate
  can give a rough CI quickly.
