---
title: "Multistart estimation"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Multistart estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

# The question multistart answers

A population model is fit by minimising an objective function, and the
minimiser only ever finds *a* minimum -- the one downhill from wherever it
started.  Nothing in a converged fit tells you whether a different set of
initial estimates would have found somewhere better.

`multistart()` answers that directly: it re-estimates the model from many
perturbed starting points and shows you what each one found.  If every start
lands on the same objective function, the fit is at a well-identified optimum.
If several starts settle at visibly different values, the fit you had was one
of several local optima and the parameter estimates should not be trusted at
face value.

```{r message = FALSE}
library(nlmixr2extra)
```

# A first run

Start from an ordinary fit.

```{r message = FALSE, warning = FALSE, results = "hide"}
one.compartment <- function() {
  ini({
    tka <- 0.45
    tcl <- 1
    tv <- 3.45
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v <- exp(tv)
    linCmt() ~ add(add.sd)
  })
}

fit <- nlmixr2(one.compartment, nlmixr2data::theo_sd, est = "focei",
               control = list(print = 0))
```

Then hand that fit to `multistart()`.  Every option lives in
`multistartControl()`, which `multistart()` accepts as a plain list.

```{r eval = FALSE}
ms <- multistart(fit, control = list(n = 20, spread = 0.3))
ms
```

The first candidate is always the unperturbed starting point, so the fit you
began with is always in the comparison.  The other `n - 1` candidates perturb
every unfixed population parameter on the scale it is estimated on -- for a
mu-referenced parameter that is usually the log scale -- and clip the result to
the parameter's declared bounds.  Between-subject variances are perturbed
multiplicatively, so they stay positive.

`ms$summary` has one row per estimated start, sorted best first, with the
objective function, its distance from the best (`dOBJF`), the information
criteria, whether the start converged, whether it finished with a parameter at
a boundary, and the final estimate of every parameter.  `ms$best` is the best
fit found and behaves like any other fit.

# Reading the two plots

```{r eval = FALSE}
plot(ms)
```

The waterfall plot puts the starts in order, best to worst, and draws each
one's objective function relative to the best.  What you want to see is a flat
run of bars at zero: every start found the same optimum.  A staircase means
several optima; the height of each step is the objective-function penalty for
landing there.  Bars are coloured by whether the start converged and whether it
finished at a boundary, and the subtitle counts the starts that failed
outright.

```{r eval = FALSE}
plot(ms, "parameters")
```

The parameter-stability plot takes the best `kBest` starts and shows how each
parameter estimate varies across them, with the best start's value as a dashed
reference line.  This is the more diagnostic of the two: a parameter whose
estimate wanders while the objective function barely moves is not identified by
the data, whichever optimum you take.

# Controlling the search

## How far the starts spread

`spread` is the fractional half-width of the perturbation, so `spread = 0.3`
moves a parameter within 30% of its value.  `omegaFold` does the same job for
the variances, as a fold-range: the default of `2` draws each variance between
half and twice its initial value.

`sampling` picks how the interval is covered.  `"uniform"` is the default;
`"lhs"` uses a Latin hypercube, which covers the range more evenly for the same
number of starts and is usually the better choice when `n` is small; `"normal"`
concentrates the starts near the initial estimates with occasional far
excursions.

`which` restricts the perturbation to named parameters, which is useful when
you suspect one particular part of the model:

```{r eval = FALSE}
ms <- multistart(fit, control = list(n = 20, which = c("tka", "tcl")))
```

## Screening

Fully estimating twenty starts is expensive, and many of them start somewhere
hopeless.  By default `multistart()` first evaluates each candidate with an
empirical-Bayes step only -- a small fraction of the cost of an estimation --
and fully estimates just the best `nFit`:

```{r eval = FALSE}
ms <- multistart(fit, control = list(n = 50, nFit = 10, spread = 0.5))
```

`ms$starts` records the screening objective for every candidate and which of
them were estimated.  Pass `screen = "none"` to estimate all of them.

Screening is a cheap approximation, so it is a way of spending a fixed budget
well rather than a guarantee: a candidate that screens badly can still be the
one that would have found the best optimum.  When the budget allows it,
`screen = "none"` is the more thorough answer.

## Where to perturb around

Starting from a fit, `around = "final"` (the default) perturbs around that
fit's final estimates, which asks "is this a local optimum?".  `around =
"initial"` perturbs around the estimates the fit started from, which asks "how
sensitive was this fit to where I started it?".  Both are worth running.

You can also skip the first fit entirely and give `multistart()` a model and
data:

```{r eval = FALSE}
ms <- multistart(one.compartment, nlmixr2data::theo_sd,
                 est = "focei", control = list(n = 20))
```

# Long runs

Each start is cached to disk as it completes, so an interrupted run resumes
where it stopped -- just call `multistart()` again with the same arguments.
Asking for more starts re-uses the ones already estimated:

```{r eval = FALSE}
ms <- multistart(fit, control = list(n = 10))   # estimates 10
ms <- multistart(fit, control = list(n = 30))   # estimates the 20 new ones
```

Anything that changes what a starting point *is* -- `sampling`, `spread`,
`which`, `perturbOmega`, `omegaFold` or `seed` -- gives the run its own cache,
so a cached estimation is never reported against a start it did not come from.
Pass `restart = TRUE` to discard the cache, or `cacheDir = NA` to never write
one.

`cores` estimates several starts at once.  Be careful with it: each estimation
already runs across every available thread, so `multistart()` pins each worker
to a single thread to avoid oversubscribing the machine.  Whether that is
faster than the serial default depends on how well the model parallelises over
subjects -- a model that already uses all your cores efficiently is usually
better left serial.  Parallel estimation is not available on Windows.

# Reproducibility

Every nlmixr2 estimator runs inside its own seeded block, so calling
`set.seed()` before a fit does not change it.  `multistart()` follows the same
discipline: `multistartControl(seed=)` seeds the perturbation draws and gives
each start its own derived seed, and the run neither depends on nor disturbs
the surrounding RNG state.  The same `seed` gives the same starts and the same
results.

`ms$starts` records the seed used for each start alongside its starting
estimates, so any individual start can be reproduced on its own.
