---
title: "nlmixr2 Algebraic Solutions with Formula"
author: "Bill Denney"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{nlmixr2 Algebraic Solutions with Formula}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

Some models are able to be described by simple algebraic solutions.  To simplify
algebraic models, you can use a formula interface similar to the formula used
with the `lme4` library.  (The formula interface is described in more detail
below; knowledge of `lme4` is not required to use the formula interface.)

# Quick start

The simplest, non-trivial model is a linear model, `y = m*x + b`.  We will
generate the data for this model and then fit it to show the simplest use case
of the nlmixr2 formula interface.

```{r quickstart-setup, class.source = 'fold-show'}
library(nlmixr2extra)
library(dplyr)
library(ggplot2)
```

```{r quickstart-data-gen}
# Simulate the equation y = 3*x + 5 with normally-distributed residual error
# having a standard deviation of 5
withr::with_seed(
  5, # standardize the random seed so that the results are reproducible
  dSim <-
    data.frame(
      x = 1:100,
      y =
        (1:100)*3 +
        5 +
        rnorm(n = 100, mean = 0, sd = 5)
    )
)

ggplot(dSim, aes(x=x, y=y)) + geom_point()
```

To fit this model requires only one line of R code:

```{r quickstart-fit, class.source = 'fold-show'}
mod <-
  nlmixrFormula(
    y~m*x + b,
    data = dSim,
    start = c(m = 2.5, b = 4, addSd = 2),
    est = "focei"
  )
```

You can also use a call to `nlmixr` or `nlmixr2`, the arguments are
the same as `nlmixrFormula`:


```{r quickstart-fit2, class.source = 'fold-show'}
mod <-
  nlmixr(
    y ~ m*x + b,
    data = dSim,
    start = c(m = 2.5, b = 4, addSd = 2),
    est = "focei"
  )
```

## Quick start, mixed-effects model

A more typical use case for nlmixr2 is a mixed-effects model.  It adds
(subject-level) random effects to the model.

```{r quickstart-data-gen-nlme}
# Setup the dataset for nonlinear mixed-effects fitting

# Simulate the equation y = 3*x + 5 with normally-distributed residual error
# having a standard deviation of 5
withr::with_seed(
  5, # standardize the random seed so that the results are reproducible
  {
    dSimSetup <-
      data.frame(
        id = rep(1:10, each=10),
        x = rep(1:10, 10),
        y = 1
      )
    dSimNlmePrep <-
      nlmixrFormula(
        y~m*x + b + bRe ~ bRe|id,
        start = c(m=3, b=5, addSd=5),
        data = dSimSetup,
        est = "rxSolve"
      )
  }
)

# Modify the simulated results to be ready for fitting
dSimNlme <-
  dSimNlmePrep |>
  as.data.frame() |>
  select(id, time, x, y=sim)

ggplot(dSimNlme, aes(x=x, y=y, colour=factor(id))) + geom_point()
```

```{r quickstart-fit-nlme, class.source = 'fold-show'}
mod <-
  nlmixr(
    y~m*x + b + bRe ~ bRe|id,
    start = c(m=3, b=5, addSd=5),
    data = dSimNlme,
    est = "focei"
  )

mod
```

In this model, the fixed effects of `m` and `b` were estimated along with the
random effect of `bRe`.

## Quick start, automatic parameter fixed effects

```{r quickstart-data-gen-param}
# Setup the dataset for nonlinear mixed-effects fitting with different types of
# parameters.

# Simulate the equation y = 3*x + 4(when z = 'a') or 6(when z = 'b') with normally-distributed residual error
# having a standard deviation of 5
withr::with_seed(
  5, # standardize the random seed so that the results are reproducible
  {
    dSimSetup <-
      data.frame(
        id = rep(1:10, each=10),
        x = rep(1:10, 10),
        y = 1,
        z = sample(factor(c("a", "b")), size = 100, replace = TRUE)
      )
    # Note that we need to give `start` as a list so that `b` can carry one
    # starting value per factor level of `z` (named-vector `c()` cannot hold
    # multi-element entries).
    dSimNlmePrep <-
      nlmixrFormula(
        y~m*x + b + bRe ~ bRe|id,
        start = list(m=3, b=c(4, 2), addSd=5),
        param = list(b ~ z),
        data = dSimSetup,
        est = "rxSolve"
      )
  }
)

# Modify the simulated results to be ready for fitting
dSimNlme <-
  dSimNlmePrep |>
  as.data.frame() |>
  select(id, time, x, y=sim, z)

ggplot(dSimNlme, aes(x=x, y=y, colour=z)) +
  geom_point() +
  geom_line(aes(group = id), colour = "gray")
```

```{r quickstart-fit-param, class.source = 'fold-show'}
mod <-
  nlmixr(
    y~m*x + b + bRe ~ bRe|id,
    start = list(m=3, b=c(4, 2), addSd=5),
    param = list(b ~ z),
    data = dSimNlme,
    est = "focei"
  )

mod
```

## Quick start, continuous covariate on a parameter

`param` also accepts numeric (continuous) covariate columns. When `b` is
modelled as a linear function of a continuous covariate `w`, two parameters
are introduced: `pop.b` (the intercept) and `cov_w_b` (the slope on `w`). The
generated model line is `b <- pop.b + cov_w_b * w`.

`start[["b"]]` may be a single value (treated as the intercept; the slope
starts at 0) or `c(intercept, slope)`.

```{r quickstart-data-gen-cont}
withr::with_seed(
  5,
  {
    dSimSetup <-
      data.frame(
        id = rep(1:10, each=10),
        x = rep(1:10, 10),
        y = 1,
        w = runif(100, min = 0, max = 10)
      )
    dSimContPrep <-
      nlmixrFormula(
        y~m*x + b + bRe ~ bRe|id,
        start = list(m=3, b=c(4, 0.2), addSd=5),
        param = list(b ~ w),
        data = dSimSetup,
        est = "rxSolve"
      )
  }
)

dSimCont <-
  dSimContPrep |>
  as.data.frame() |>
  select(id, time, x, w, y=sim)
```

```{r quickstart-fit-cont, class.source = 'fold-show'}
mod <-
  nlmixr(
    y~m*x + b + bRe ~ bRe|id,
    start = list(m=3, b=c(4, 0.2), addSd=5),
    param = list(b ~ w),
    data = dSimCont,
    est = "focei"
  )

mod
```

## Quick start, log-linked parameter

Some parameters are naturally strictly positive (clearances, volumes, rate
constants). Pass `paramLink = c(<parameter> = "log")` to wrap the linear
combination in `exp()`. The starting values are then on the log scale.

```{r quickstart-fit-log, class.source = 'fold-show', eval = FALSE}
nlmixr(
  y~m*x + b + bRe ~ bRe|id,
  start = list(m=3, b=c(log(4), 0.2), addSd=5),
  param = list(b ~ w),
  paramLink = c(b = "log"),
  data = dSimCont,
  est = "focei"
)
```

## Worked example: concentration-QT analysis

A common use of the formula interface is a thorough QT (TQT) or
concentration-QT (C-QT) analysis. The clinical question is whether a 10 msec
$\Delta QT$ is exceeded at therapeutic drug concentrations. The standard
model has:

- A categorical effect for nominal post-dose time (captures circadian and
  food-related drift)
- A continuous linear effect for drug concentration (the parameter of
  interest)
- A per-subject random intercept (baseline $\Delta QT$ offset)

In the formula interface this is one call:

```
dQT ~ b + bRe ~ (bRe|id)
param  = list(b ~ timeF + conc)
```

`b` is a single intercept parameter; `param` decomposes it into a fixed
effect per `timeF` level *plus* a linear slope on `conc`. The random per-
subject offset `bRe` is added to `b` in the predictor.

### Simulating the data

We start from a one-compartment oral PK profile (Bateman equation), scale it
so the typical $C_{max}$ is about 1000 ng/mL, and apply per-subject
log-normal scaling so individual $C_{max}$ values are log-normally
distributed.

```{r cqt-data-gen, class.source = 'fold-show'}
withr::with_seed(42, {
  nSubj    <- 40
  nomTimes <- c(0, 0.5, 1, 1.5, 2, 3, 4, 6, 8, 12)  # 10 nominal times

  # Typical 1-compartment oral PK profile (unit Bateman), then scaled
  # so the population Cmax is 1000 ng/mL
  bateman    <- function(t, ka = 2, ke = 0.1) {
    (ka / (ka - ke)) * (exp(-ke * t) - exp(-ka * t))
  }
  cmaxFactor <- 1000 / max(bateman(seq(0, 12, 0.01)))

  # Per-subject log-normal scaling of the concentration profile
  subjScale  <- exp(rnorm(nSubj, mean = 0, sd = 0.4))

  cqtDesign <-
    expand.grid(id = 1:nSubj, timeH = nomTimes)
  cqtDesign$conc  <- bateman(cqtDesign$timeH) * cmaxFactor *
                     subjScale[cqtDesign$id]
  cqtDesign$timeF <- factor(cqtDesign$timeH,
                            levels = as.character(nomTimes))
  cqtDesign$dQT   <- 1  # placeholder; nlmixrFormula(est="rxSolve") fills it in
})
```

Now we simulate `dQT` from the true model. The true slope is 0.01
msec/(ng/mL), which produces about a 10 msec $\Delta QT$ at the population
$C_{max}$. The categorical time effects encode a typical circadian drift
pattern. The simulation's `bRe` SD is 1 (the formula interface's current
default for random-effect starting values).

```{r cqt-sim, class.source = 'fold-show'}
cqtSimPrep <-
  nlmixrFormula(
    dQT ~ b + bRe ~ (bRe|id),
    data  = cqtDesign,
    # 10 time-level intercepts (relative to time 0) + slope on conc
    start = list(
      b     = c(0,         # timeF == 0  (pre-dose baseline)
                3, 5, 6, 5, 2, 0, -1, -2, -3,  # post-dose time effects
                0.01),     # slope on conc (msec per ng/mL)
      addSd = 3
    ),
    param = list(b ~ timeF + conc),
    est   = "rxSolve"
  )

cqtData <-
  cqtSimPrep |>
  as.data.frame() |>
  select(id, conc, timeF, dQT = sim)

ggplot(cqtData, aes(x = conc, y = dQT, colour = timeF)) +
  geom_point(alpha = 0.7) +
  labs(x = "Concentration (ng/mL)", y = expression(Delta * "QT (msec)"))
```

### Fitting the model

Algebraic models with both a factor and a continuous covariate on the same
parameter can be hard for `focei`'s gradient calculation when there are
many factor levels. A robust two-stage strategy is to fit the fixed effects
with `bobyqa` (which is gradient-free and converges fast for this class of
model) and then refine with `focei` using the bobyqa estimates as starts,
which adds the random effect.

```{r cqt-fit-fe, class.source = 'fold-show'}
fitFE <-
  nlmixr(
    dQT ~ b,                                       # fixed effects only
    data  = cqtData,
    start = list(
      b     = c(0, 3, 5, 6, 5, 2, 0, -1, -2, -3, 0.01),
      addSd = 3
    ),
    param = list(b ~ timeF + conc),
    est   = "bobyqa"
  )

fitFE
```

```{r cqt-fit-re, class.source = 'fold-show'}
# Carry the bobyqa estimates forward as starts for focei + random effects
feEst <- setNames(fitFE$iniDf$est, fitFE$iniDf$name)

fitRE <-
  nlmixr(
    dQT ~ b + bRe ~ (bRe|id),
    data  = cqtData,
    start = list(
      b     = c(feEst[1:10], feEst[["cov_conc_b"]]),
      addSd = feEst[["addSd"]]
    ),
    param = list(b ~ timeF + conc),
    est   = "focei"
  )

fitRE
```

The recovered `cov_conc_b` estimates the slope of concentration on
$\Delta QT$. Multiplying by the population $C_{max}$ (about 1000 ng/mL)
gives the projected $\Delta QT$ at $C_{max}$ — typically within ~1 msec of
the simulated 10 msec.

# Defining the model equation (the formula)

The model formula has 3 parts: the dependent variable, the equation predicting
the dependent variable, and the optional random effects.  A full equation looks
like the below:

```
dv~predictors~randomEffects
```

Any model part may have any parameter name (for example, the dependent variable
does not have to be named `dv`).

## Dependent variable

The dependent variable may be any column in the `data`.  It must only be the
column name and no additional modifiers.  For example, `log(param)` is not
allowed.

## Predictors

Predictors may be any algebraic equation.  In general, there are no restrictions
to the way that the equation may be written.

## Random effects

Random effects are not required for a model; fixed-effect-only models are
possible.  When present, fixed and random effects must have different names.

Random effects are written as a parameter a pipe (`|`) and its grouping value.
From the quick start example above, the parameter is `bRe` and the grouping
value is `id`, so it is written `bRe|id`.

For multiple parameters, they must be setup separately.  For example, to have
`mRe` and `bRe` both grouped by `id`, it would be defined as
`~(mRe|id)+(bRe|id)`.

Only one grouping variable is supported per model. Multiple grouping
variables (for inter-occasion variability, for example) are not yet
supported.

## Per-parameter covariate models (`param`)

The `param` argument adds a covariate model to one or more parameters. It
accepts a formula or a list of formulas, each of the form
`<parameter> ~ <covariate columns>`. The right-hand side names one or more
columns from `data`:

- A **factor** column produces one fixed effect per level. The first level is
  the baseline; subsequent levels are estimated as differences from the
  baseline.
- A **numeric** column produces a slope. With only the continuous covariate,
  two parameters are introduced — `pop.<parameter>` (intercept) and
  `cov_<column>_<parameter>` (slope).
- Multiple covariates can be combined on the same parameter with `+`:
  `b ~ z + w` combines factor `z` and continuous `w`. When a factor is
  present, the factor supplies the baseline so `pop.b` is not introduced.

When mixing factor and continuous covariates on the same parameter,
`start[[<parameter>]]` must explicitly list one value per factor level
followed by one slope per continuous covariate (in the order they appear in
`param`).

### `paramLink` (link function per parameter)

By default the assembled linear combination is the parameter itself
(`b <- <linear combination>`). Set `paramLink = c(b = "log")` to wrap the
combination in `exp()`, so the parameter is strictly positive on its natural
scale.

# Providing the Starting estimates for the model

## Fixed effects

Fixed effects are defined by the `start` argument.  The `start` argument may
either be a named vector or a named list.  If fixed effects only have a single
starting value, then the two methods are equivalent.  `c(m=3, b=5)` is the same
as `list(m=3, b=5)`.  If one of the fixed effects is defined by a factor
variable (more on that later), the list may have multiple starting values such
as `list(m=3, b=c(5, 10))`.

For continuous covariates passed via `param`, `start[[<parameter>]]` may be a
single intercept value (the slope defaults to 0) or `c(intercept, slope)`.

## Random effects

Random effect starting values are currently fixed at 1 without the ability to
modify it.

## Residual error models

The default residual error model is `~ add(addSd)` (additive standard
deviation `addSd`). More complex specifications are supported as long as
every sigma parameter in the residual formula appears in `start`. For
example:

```r
nlmixr(
  y~m*x + b ~ (bRe|id),
  start = list(m=3, b=5, addSd=1, propSd=0.1),
  residualModel = ~ add(addSd) + prop(propSd),
  data = dSimNlme,
  est = "focei"
)
```
