---
title: "Stabilizing Covariance Estimates with preconditionFit()"
author: "Matthew Fidler"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Stabilizing Covariance Estimates with preconditionFit()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# The problem: ill-conditioned covariance matrices

After fitting a nonlinear mixed-effects model with FOCEI, the standard
errors (SEs) of the population parameters are derived from the
variance-covariance matrix of the parameter estimates. This matrix is
assembled from the **R matrix** (second-derivative / Hessian approximation)
and, optionally, the **S matrix** (cross-product of first derivatives). When
the parameters are on very different scales, or when two parameters are
nearly collinear, the R matrix becomes ill-conditioned: small rounding
errors in the gradient calculations are amplified during the matrix inversion
and the resulting covariance matrix is unreliable.

Symptoms of this problem include:

- `covMethod` reported as `"r"` or `"s"` rather than the preferred `"r,s"`
- NaN or negative values on the diagonal of the covariance matrix
- Standard errors that are implausibly large or small relative to the estimates
- `%RSE` values that are very large

# The solution: preconditioning

Aoki, Nordgren, and Hooker (2016) showed that a simple **linear
reparameterization** of the fixed-effects parameters can dramatically improve
the condition number of the R matrix without changing the model or its fit. The
idea is to find a matrix **P** such that **P R P^T** is close to the identity
matrix, run FOCEI on the reparameterized model (which converges to a
numerically stable covariance matrix), and then transform the result back to
the original parameterization.

`nlmixr2extra` implements this approach in `preconditionFit()`.

> **Reference:** Aoki Y, Nordgren R, Hooker AC. Preconditioning of Nonlinear
> Mixed Effects Models for Stabilization of Variance-Covariance Matrix
> Computations. *AAPS J.* 2016;18(2):505--518.
> doi:[10.1208/s12248-016-9866-5](https://doi.org/10.1208/s12248-016-9866-5)

# Basic usage

```{r basic, eval=FALSE}
library(nlmixr2extra)

oneCompartment <- function() {
  ini({
    tka <- 0.45  # log Ka
    tcl <- 1     # log Cl
    tv  <- 3.45  # log V
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v  ~ 0.1
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv  + eta.v)
    d/dt(depot)  <- -ka * depot
    d/dt(center) <-  ka * depot - cl/v * center
    cp <- center / v
    cp ~ add(add.sd)
  })
}

fit <- nlmixr2(oneCompartment, data = nlmixr2data::theo_sd,
               est = "focei", control = list(print = 0))
fit
```

If the covariance method reported is not `"r,s"` (or you simply want a more
robust covariance estimate), apply preconditioning:

```{r precondition-basic, eval=FALSE}
preconditionFit(fit)
```

`preconditionFit()` modifies the **original fit object in-place**: after the
call, `fit$cov`, `fit$parFixedDf`, and `fit$covMethod` all reflect the
preconditioned covariance. The function returns the preconditioned covariance
matrix invisibly.

# Controlling the amount of re-estimation

The `estType` argument determines how much of the model is re-estimated on the
reparameterized problem:

| `estType` | What is re-estimated | When to use |
|-----------|----------------------|-------------|
| `"full"` (default) | Outer *and* inner iterations (theta, ETA) | When you suspect the original estimates may have converged to a local minimum |
| `"posthoc"` | Inner iterations only (ETA, given theta) | When you trust the theta estimates but want a better covariance |
| `"none"` | Nothing — only the covariance matrix is re-computed | Fastest; good for a quick diagnostic check |

```{r esttypes, eval=FALSE}
# Only recompute the covariance matrix (fastest)
preconditionFit(fit, estType = "none")

# Fix theta, re-estimate ETAs, then compute covariance
preconditionFit(fit, estType = "posthoc")

# Full re-estimation on the reparameterized model (default)
preconditionFit(fit, estType = "full")
```

# Retrying until convergence: `ntry`

The inner loop of `preconditionFit()` iterates the preconditioning until
the reparameterized problem achieves a `"r,s"` covariance (the most reliable
type). If that target is not reached within `ntry` attempts the function
stops with an error. The default is `ntry = 10`.

```{r ntry, eval=FALSE}
# Allow up to 20 attempts before giving up
preconditionFit(fit, ntry = 20L)
```

If `preconditionFit()` fails even with more tries, the R matrix itself may be
too poorly determined for preconditioning to help. In that case inspect the
model for identifiability problems or consider likelihood profiling
(`profile(fit)`) instead of Wald-based confidence intervals.

# Switching between covariance estimates

`preconditionFit()` stores the preconditioned covariance alongside any
previously computed covariances in `fit$covList`. You can inspect what is
available and switch between them with `setCov()`:

```{r setcov, eval=FALSE}
# See which covariance estimates are stored
names(fit$covList)
# e.g. c("r,s", "precondition")

# Switch back to the standard r,s covariance
setCov(fit, "r,s")

# Switch to the preconditioned covariance
setCov(fit, "precondition")
```

After `setCov()` the fit object is updated in-place and the displayed
parameter table (`fit$parFixedDf`) reflects the selected covariance.

# Worked example: comparing SEs before and after preconditioning

```{r compare, eval=FALSE}
library(nlmixr2extra)

fit <- nlmixr2(oneCompartment, data = nlmixr2data::theo_sd,
               est = "focei", control = list(print = 0))

# Record the original parameter table
dfBefore <- fit$parFixedDf
cat("Covariance method before:", fit$covMethod, "\n")

# Apply preconditioning (only recompute covariance, do not re-estimate)
preconditionFit(fit, estType = "none")
dfAfter <- fit$parFixedDf
cat("Covariance method after:", fit$covMethod, "\n")

# Compare standard errors
cbind(
  SE_before = dfBefore$SE,
  SE_after  = dfAfter$SE,
  row.names = rownames(dfBefore)
)
```

The point estimates (`Estimate`, `Back-transformed`) and the random-effects
summaries (`BSV(CV%)`, `Shrink(SD)%`) are identical before and after — only
the covariance-derived quantities (SE, %RSE, confidence intervals) change.

# When to use `preconditionFit()`

Use `preconditionFit()` when:

- `fit$covMethod` is not `"r,s"` after a FOCEI fit
- Standard errors look implausibly large or contain NaN
- You want a publication-quality covariance estimate and are willing to spend
  extra computation time

It is not needed when FOCEI already produces a `"r,s"` covariance and the SEs
look reasonable. For models where identifiability is in doubt (very large
%RSE for multiple parameters simultaneously), consider likelihood profiling
rather than trying to stabilize a structurally unreliable covariance.

# Requirements

`preconditionFit()` requires:

- A fit produced by FOCEI (or a method that stores an R matrix in `fit$R`).
  SAEM fits do not store an R matrix by default; call
  `fit <- nlmixr2(model, data, est = "focei", ...)` first, or use
  `getVarCov(saemFit)` to trigger covariance computation.
- The `symengine` package (used to construct the symbolic reparameterization).
- The `nlmixr2extra` package to be loaded.
